---
title: "Spectral Scatterplot of Sentinel-2 Pixels"
author: "Iris Nana Obeng"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  markdown: 
    wrap: 70
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This project uses Sentinel-2B multispectral imagery to explore the spectral
characteristics of pixels within a selected 4 km × 4 km study area.

The analysis uses four 10 m Sentinel-2 bands:

* **B02 (Blue, ~490 nm)** — used on the x-axis and for clustering
* **B03 (Green, ~560 nm)** — used for clustering
* **B04 (Red, ~665 nm)** — used for clustering
* **B08 (Near-infrared, ~842 nm)** — used on the y-axis, and deliberately
  excluded from the clustering so that the visualisation shows the clusters
  in a band the classifier did not use.

An unsupervised k-means classification is applied to the Blue, Green, and Red
reflectance bands. The resulting clusters are then visualised using a
Blue–NIR scatterplot.

The final figure follows the marginal-density approach described by
Claus O. Wilke in *Fundamentals of Data Visualization*. Density curves are
placed directly above and to the right of the scatterplot. The top density
is directly labelled with the cluster identities, replacing the need for a
conventional legend.

# Packages

```{r packages, message=FALSE, warning=FALSE}
library(terra)
library(ggplot2)
library(dplyr)
library(imageRy)
library(cowplot)
```

# Load Sentinel-2 bands

The Sentinel-2B tile used in this analysis is T30NXN, acquired on
26 January 2022.

```{r load-bands}
project_root <- "/Users/irisnanaobeng/Desktop/Scatter Plot Algorithim/Scatter_Plot_Algorithim"

blue  <- rast(file.path(project_root, "T30NXN_20220126T102209_B02_10m.jp2"))
green <- rast(file.path(project_root, "T30NXN_20220126T102209_B03_10m.jp2"))
red   <- rast(file.path(project_root, "T30NXN_20220126T102209_B04_10m.jp2"))
nir   <- rast(file.path(project_root, "T30NXN_20220126T102209_B08_10m.jp2"))
```

# Combine bands

```{r combine-bands}
sentinel_stack <- c(blue, green, red, nir)

names(sentinel_stack) <- c(
  "Blue",
  "Green",
  "Red",
  "NIR"
)
```

# The study area

The analysis uses a 4 km × 4 km area within the Sentinel-2 tile. The crop
begins approximately 60 km from the lower-left corner of the tile in both
the x and y directions.

```{r crop}
crop_extent <- ext(
  xmin(sentinel_stack) + 60000,
  xmin(sentinel_stack) + 64000,
  ymin(sentinel_stack) + 60000,
  ymin(sentinel_stack) + 64000
)

sentinel_crop <- crop(sentinel_stack, crop_extent)
```

# Scale reflectance

Sentinel-2 reflectance values are scaled by 10,000. They are converted here
to approximately 0–1 reflectance values.

```{r scale-reflectance}
sentinel_crop <- sentinel_crop / 10000
```

# The clustering image

The clustering uses the Blue, Green and Red bands. NIR is deliberately
excluded so it can be used as the y-axis of the scatterplot.

```{r clustering-stack}
rgb_stack <- sentinel_crop[[c("Blue", "Green", "Red")]]
```

# K-means clustering

```{r classification}
set.seed(42)

classified <- imageRy::im.classify(
  rgb_stack,
  num_clusters = 3,
  seed = 42,
  do_plot = FALSE
)
```

The resulting clusters represent groups of pixels with similar values in the
input bands. The numerical cluster IDs do not have an inherent land-cover
meaning.

# Pixel data

```{r pixel-data}
pixel_data <- data.frame(
  Blue    = as.numeric(values(sentinel_crop[["Blue"]],  mat = FALSE)),
  NIR     = as.numeric(values(sentinel_crop[["NIR"]],   mat = FALSE)),
  Cluster = as.numeric(values(classified,               mat = FALSE))
)

pixel_data <- pixel_data |>
  filter(
    !is.na(Blue),
    !is.na(NIR),
    !is.na(Cluster)
  )
```

# Descriptive cluster labels

The numerical cluster IDs produced by k-means are arbitrary. To make the
figure easier to interpret, the clusters are renamed A, B and C according to
their abundance.

```{r cluster-labels}
cluster_sizes <- pixel_data |>
  count(Cluster, sort = TRUE)

cluster_labels <- cluster_sizes$Cluster

pixel_data <- pixel_data |>
  mutate(
    Cluster_Label = case_when(
      Cluster == cluster_labels[1] ~ "A",
      Cluster == cluster_labels[2] ~ "B",
      Cluster == cluster_labels[3] ~ "C"
    )
  )
```

The labels A, B and C are descriptive identifiers only. They are not
validated land-cover classes.

# Sample pixels for the scatterplot

The full dataset contains approximately 160,000 10 m pixels. To keep the
central scatterplot readable, 300 pixels are randomly sampled for display.

The sampling affects the scatterplot only. The classification and the
marginal density distributions use the full classified dataset.

```{r sample-pixels}
set.seed(1)

plot_data <- pixel_data |>
  slice_sample(n = 300)
```

# Colours and shapes

```{r palette}
cluster_colours <- c(
  A = "#011959",
  B = "#D68A00",
  C = "#A3311D"
)

cluster_shapes <- c(
  A = 16,
  B = 17,
  C = 15
)
```

# Central scatterplot

```{r main-scatter}
main_plot <- ggplot(
  plot_data,
  aes(
    x = Blue,
    y = NIR
  )
) +

  geom_point(
    aes(
      colour = Cluster_Label,
      shape  = Cluster_Label
    ),
    size  = 1.8,
    alpha = 0.6
  ) +

  scale_colour_manual(
    values = cluster_colours
  ) +

  scale_shape_manual(
    values = cluster_shapes
  ) +

  labs(
    title    = "Spectral Scatterplot of Classified Sentinel-2 Pixels",
    subtitle = "300 sampled pixels with marginal cluster distributions",
    x        = "Blue reflectance (B02)",
    y        = "Near-infrared reflectance (B08)"
  ) +

  theme_minimal(
    base_size = 13
  ) +

  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(size = 11)
  )
```

# Directly labelled marginal densities

The marginal densities follow the direct-labelling approach described by
Wilke. The density curves are calculated separately for each cluster and
then inserted into the margins of the scatterplot.

```{r wilke-marginals, fig.width=9, fig.height=7}


blue_dens <- pixel_data |>
  group_by(Cluster_Label) |>
  group_modify(~ {
    d <- density(.x$Blue, na.rm = TRUE)
    data.frame(Blue = d$x, density = d$y)
  }) |>
  ungroup()



nir_dens <- pixel_data |>
  group_by(Cluster_Label) |>
  group_modify(~ {
    d <- density(.x$NIR, na.rm = TRUE)
    data.frame(NIR = d$x, density = d$y)
  }) |>
  ungroup()



blue_max <- blue_dens |>
  group_by(Cluster_Label) |>
  filter(density == max(density)) |>
  ungroup()



blue_labels <- blue_max |>
  mutate(
    label = Cluster_Label,

    hjust = case_when(
      Cluster_Label == "A" ~ 0,
      Cluster_Label == "B" ~ 0.5,
      Cluster_Label == "C" ~ 0
    ),

    vjust = case_when(
      Cluster_Label == "A" ~ 1,
      Cluster_Label == "B" ~ 0,
      Cluster_Label == "C" ~ 1
    ),

    nudge_x = case_when(
      Cluster_Label == "A" ~ 0.003,
      Cluster_Label == "B" ~ 0,
      Cluster_Label == "C" ~ 0.003
    ),

    nudge_y = case_when(
      Cluster_Label == "A" ~ 3,
      Cluster_Label == "B" ~ 5,
      Cluster_Label == "C" ~ 3
    )
  )


xdens <- axis_canvas(
  main_plot,
  axis = "x"
) +

  geom_ribbon(
    data = blue_dens,
    aes(
      x = Blue,
      ymin = 0,
      ymax = density,
      fill = Cluster_Label,
      group = Cluster_Label
    ),
    alpha = 0.35
  ) +

  geom_line(
    data = blue_dens,
    aes(
      x = Blue,
      y = density,
      colour = Cluster_Label,
      group = Cluster_Label
    ),
    linewidth = 0.5
  ) +

  geom_text(
    data = blue_labels,
    aes(
      x = Blue + nudge_x,
      y = density + nudge_y,
      label = label,
      hjust = hjust,
      vjust = vjust,
      colour = Cluster_Label
    ),
    size = 13 / .pt,
    fontface = "bold",
    inherit.aes = FALSE
  ) +

  scale_colour_manual(
    values = cluster_colours,
    guide = "none"
  ) +

  scale_fill_manual(
    values = cluster_colours,
    guide = "none"
  ) +

  scale_y_continuous(
    limits = c(0, max(blue_dens$density) * 1.10),
    expand = c(0, 0)
  ) +

  coord_cartesian(clip = "off")



ydens <- axis_canvas(
  main_plot,
  axis = "y",
  coord_flip = TRUE
) +

  geom_ribbon(
    data = nir_dens,
    aes(
      x = NIR,
      ymin = 0,
      ymax = density,
      fill = Cluster_Label,
      group = Cluster_Label
    ),
    alpha = 0.35
  ) +

  geom_line(
    data = nir_dens,
    aes(
      x = NIR,
      y = density,
      colour = Cluster_Label,
      group = Cluster_Label
    ),
    linewidth = 0.5
  ) +

  scale_colour_manual(
    values = cluster_colours,
    guide = "none"
  ) +

  scale_fill_manual(
    values = cluster_colours,
    guide = "none"
  ) +

  scale_y_continuous(
    limits = c(0, max(nir_dens$density) * 1.05),
    expand = c(0, 0)
  ) +

  coord_flip()



p1 <- insert_xaxis_grob(
  main_plot + theme(legend.position = "none"),
  xdens,
  grid::unit(3 * 14, "pt"),
  position = "top",
  clip = "off"
)


p2 <- insert_yaxis_grob(
  p1,
  ydens,
  grid::unit(3 * 14, "pt"),
  position = "right",
  clip = "off"
)



marginal_plot <- ggdraw(p2)

marginal_plot


ggsave(
  file.path(project_root, "spectral_scatterplot_marginals.png"),
  plot   = marginal_plot,
  width  = 9,
  height = 7,
  dpi    = 300
)

```


# Interpreting the scatterplot

The three clusters occupy different regions of the Blue–NIR spectral space.

The labels A, B and C are descriptive cluster identifiers rather than
validated land-cover classes. Their spectral characteristics can therefore
be used to describe patterns in the image, but they should not automatically
be interpreted as specific land-cover types.

# Interpreting the marginal densities

The **top marginal density** shows the distribution of Blue reflectance within each cluster. The labels A, B and C are positioned relative to the peaks of the corresponding density curves, allowing the reader to associate each letter with its curve without consulting a separate key.

The **right marginal density** shows the distribution of NIR reflectance within each cluster. It is deliberately left unlabelled: because the top marginal already identifies each cluster by letter, labelling the right marginal would be redundant. Colour matching is sufficient to relate the right-hand curves to the labelled curves above.

The top and right marginals use different vertical scales because their densities are computed for different variables. Curve heights should therefore not be compared directly across the two marginals.


# Limitations

1. **Unsupervised clustering**

   K-means identifies groups of spectrally similar pixels rather than
   validated land-cover classes.

2. **Limited clustering variables**

   Only B02, B03 and B04 are used to compute the clusters. Other Sentinel-2
   bands, such as SWIR, could provide additional spectral information.

3. **Different visualisation space**

   The clusters are calculated using B02, B03 and B04, while the scatterplot
   displays B02 against B08. The plotted Blue–NIR separation is therefore not
   the same as the three-band space in which k-means performed the clustering.

4. **Sampled scatterplot**

   The central scatterplot displays 300 randomly sampled pixels.

5. **Full-data marginal distributions**

   The marginal densities are calculated from the full classified dataset.

6. **No accuracy assessment**

   The clusters have not been compared with independent reference data.

# Conclusion

This analysis uses unsupervised k-means clustering to explore spectral
variation within a Sentinel-2B image.

The final figure is a single visual object: a Blue–NIR scatterplot with
marginal density distributions attached directly to its top and right
edges.

The top density carries the cluster labels A, B and C directly on the
curves. The right density identifies its clusters by colour alone. Together,
these replace the conventional legend, following Wilke's approach of
placing labels close to the graphical elements they describe.

# References

Wilke, C. O. (2019). *Fundamentals of Data Visualization: A Primer on
Making Informative and Compelling Figures*. O'Reilly Media.


European Space Agency. *Copernicus Sentinel-2 MSI Level-2A*. Tile T30NXN,
26 January 2022. 